---
title: "BIOSTAT 620 Lab 6"
author: "Kyoka Ono"
format: html
editor: visual
---

### Read in Medical Transcriptions

```{r}
library(tidyverse)
library(tidytext)
```

```{r}
url <- "https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/mtsamples/mtsamples.csv"

medical_data <- read.csv(url)
```

### Question 1: What specialties do we have?

```{r}
medical_data %>% count(medical_specialty) %>% View()
```

```{r}
medical_data %>% count(medical_specialty) %>% nrow()
```

Surgery and Neurosurgery are overlapping

### Question 2

```{r}
medical_data %>% 
  unnest_tokens(tokens, transcription) %>% 
  count(tokens) %>% 
  top_n(20,n) %>%
  ggplot(aes(n, fct_reorder(tokens, n))) + 
  geom_col(fill = "skyblue") + 
  labs(x = "Tokens Frequency", y = "Tokens")
```

The output makes sense since the top most frequent words consist of articles, prepositions, and pronouns.

### Question 3

```{r}
medical_data %>% 
  unnest_tokens(tokens, transcription) %>% 
  anti_join(stop_words, by = c("tokens" = "word")) %>% 
  count(tokens, sort = TRUE) %>%
  top_n(20,n) %>% 
  ggplot(aes(n, fct_reorder(tokens, n))) + 
  geom_col(fill = "skyblue") + 
  labs(x = "Tokens Frequency", y = "Tokens")
```

### Question 4

```{r}
# bi-grams
medical_data %>% 
  unnest_ngrams(bigrams, transcription, n = 2) %>%
  count(bigrams) %>% 
  top_n(20,n) %>%
  ggplot(aes(n, fct_reorder(bigrams, n))) + 
  geom_col(fill = "skyblue") + 
  labs(x = "Bigrams Frequency", y = "Bigrams")
```

```{r}
# tri-grams
medical_data %>% 
  unnest_ngrams(trigrams, transcription, n = 3) %>%
  count(trigrams) %>% 
  top_n(20,n) %>%
  ggplot(aes(n, fct_reorder(trigrams, n))) + 
  geom_col(fill = "skyblue") + 
  labs(x = "Trigrams Frequency", y = "Trigrams")
```

### Question 5

-   selected word: patient

```{r}
# before
medical_data %>% 
  unnest_ngrams(bigrams, transcription, n = 2) %>%
  separate(bigrams, into = c("word1", "word2"), sep = " ") %>%
  select(word1, word2) %>%
  filter(word2 == "patient") %>% count(word1, sort = TRUE)
```

```{r}
# after
medical_data %>% 
  unnest_ngrams(bigrams, transcription, n = 2) %>%
  separate(bigrams, into = c("word1", "word2"), sep = " ") %>%
  select(word1, word2) %>%
  filter(word1 == "patient") %>% count(word2, sort = TRUE)
```

### Question 6

```{r}
medical_data %>% 
  unnest_tokens(tokens, transcription) %>% 
  anti_join(stop_words, by = c("tokens" = "word")) %>%
  group_by(medical_specialty) %>% 
  count(tokens) %>% 
  top_n(5,n) 
```

### Question 7

In each medical specialty, which top two keywords appear the most?

```{r}
medical_data %>% 
  unnest_tokens(tokens, keywords) %>% 
  anti_join(stop_words, by = c("tokens" = "word")) %>%
  group_by(medical_specialty) %>%
  count(tokens) %>% 
  top_n(2,n)
```
