---
title: "BIOSTAT 620 Lab 3"
author: "Kyoka Ono"
date: "2026-01-29"
output: html_document
---

```{r}
library(dplyr)
```

### 1. Read in the data

```{r}
met <- read.csv("/Users/kyokaono/Desktop/BIOSTAT620 /met_all (1).gz")
```

### 2. Check the dimensions, headers, footers

- 2377343 rows
- 30 columns 

```{r}
dim(met)
```

### 3. Take a look at all the variables (hint: use the str funcion)

```{r}
str(met)
```
### 4. Take a closer look at the key variales (year, day, hour, temp, elev, wind.sp). Hint: use table and/or summary.

```{r}
year <- met["year"]
table(year)
```

```{r}
day <- met["day"]
table("day")
```

```{r}
hour <- met["hour"]
table(hour)
```

```{r}
summary(hour)
```

```{r}
temp <- met["temp"]
summary(temp)
```

```{r}
elev <- met["elev"]
summary(elev)
```

```{r}
wind.sp <- met["wind.sp"]
summary(wind.sp)
```

```{r}
# checking missing data for the elevation
sum(met$elev == 9999.0)
```

```{r}
# replacing 9999.0 elevation with NAs
met$elev[which(met$elev == 9999.0)] <- NA
```

```{r}
sum(met$elev == 9999.0, na.rm = TRUE)
```

```{r}
# checking the highest weather station (4113)
max(met$elev, na.rm = TRUE)
```

```{r}
# checking temperature data with -40C
sum(met$temp == -40.0, na.rm = TRUE)
```

```{r}
# Replacing -40.0 with NA 
met$temp[which(met$temp == -40.0)] <- NA
```

```{r}
sum(met$temp == -40.0, na.rm = TRUE)
```

```{r}
# the new lowest temperature
min(met$temp, na.rm = TRUE)
```

### 5. Check the data against an extrnal data source.

Validate that the total range of elevation in the data make sense. Also, we should check the suspicious -17.2C temperature value (where is it located?).

```{r}
# The range of elevation makes sense. According to Google, it is positive if it is above the sea level, and negative if it is below the sea level. 
summary(met$elev)
```

```{r}
# According to Google map, the location is Ellicott, Colorado. The average temperature during August seems to be higher than what is represented on the data. 
head(met %>% filter(temp == -17.2))
```

### 6. Calculate summary statistics 

Remember to keep the initial question in mind. We want to pick out the weather station with maximum elevation and examine its windspeed and temperature. 

Some ideas: select the weather station with maximum elevation; within this station, look at the correlation between temperature and wind speed. Also, look at the correlation between (temp and wind.sp) and (hour and day) (four total pairwise correlations).

```{r}
max_elev_data <- met %>% filter(elev == 4113, !is.na(temp), !is.na(wind.sp))
head(max_elev_data)
```

```{r}
# temperature and windspeed 
cor(max_elev_data$temp, max_elev_data$wind.sp)
```

```{r}
# temperature and hour
cor(max_elev_data$temp, max_elev_data$hour)
```

```{r}
# temperature and day
cor(max_elev_data$temp, max_elev_data$day)
```

```{r}
# wind.sp and hour
cor(max_elev_data$wind.sp, max_elev_data$hour)
```

```{r}
# wind.sp and day
cor(max_elev_data$wind.sp, max_elev_data$day)
```

### 7. Exploratory graphs

We should look at the distributions of all of the key variables to make sure there are no remaining issues with the data. Hint: using the hist function, make histograms of the elev, temp, and wind.sp variables.

(in hist(), it ignores NA values)

```{r}
hist(met$elev)
```

```{r}
hist(met$temp)
```

```{r}
hist(met$wind.sp)
```

```{r}
# Look at where the weather station with highest elevation is located
library(leaflet)
leaflet(max_elev_data) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~lat,lng=~lon, opacity=1, fillOpacity=1, radius=100)
```

```{r}
# Look at the time series of temperature and wind speed at this location. For this we will need to create a date-time variable for the x-axis 
library(lubridate)
max_elev_data$date <- with(max_elev_data, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(max_elev_data$date)
max_elev_data <- max_elev_data[order(max_elev_data$date), ]
head(max_elev_data)
```

### 8. Ask questions

What is the correlation between wind.sp and elevation? 

```{r}
new_data <- met %>% filter(!is.na(elev), !is.na(wind.sp))
cor(new_data$elev, new_data$wind.sp)
```

Ans. There is a subtle positive correlation. 

