---
title: "BIOSTAT 620 Lab 9"
author: "Kyoka Ono"
format: html
editor: visual
embed-resources: true
---

## Intro

In this lab, you will be working with a hospital database. There will be four tables that we will be working with: **doctor**, **visit**, and **payment**. The main goal of the lab is to gain experience working with SQL databases. First, we will set up our SQL database:

```{r}
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")

# Copy data.frames to database
dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```

```{r}
dbGetQuery(con, 
           "PRAGMA table_info(doctor)")
```

## Exercise 1

Edit the code below to retrieve the doctor ID, first name and last name for all doctors using the `doctor` table. Sort by last name and then by first name.

```{r}
head(dbGetQuery(con, 
           "SELECT doctor_id, first_name, last_name
           FROM doctor
           ORDER by last_name, first_name"), 10)
```

## Exercise 2

Retrieve the doctor ID, first name, and last name for doctors whose last name equals WILLIAMS or DAVIS

```{r}
dbGetQuery(con, 
           "SELECT doctor_id, first_name, last_name
           FROM doctor
           WHERE last_name IN ('WILLIAMS', 'DAVIS')")
```

## Exercise 3

Write a query against the `visit` table that returns the IDs of the patients who visited a hospital starting on July 5, 2005 (use the visit_start_date column, and you can use the date() function to ignore the time component).

Include a single row for each distinct patient ID.

```{r}
dbGetQuery(con, 
           "PRAGMA table_info(visit)")
```

```{r}
head(dbGetQuery(con, 
           "SELECT DISTINCT patient_id
           FROM visit 
           WHERE date(visit_start_date) = '2005-07-05'"), 10)
```

## Exercise 4

```{r}
dbGetQuery(con, 
           "PRAGMA table_info(payment)")
```

### Exercise 4.1

Construct a query that retrieves all columns of rows from the `payment` table where the amount is either 199, 799, or 999.

```{r}
head(dbGetQuery(con, 
           "SELECT *
           FROM payment
           WHERE amount IN (199, 799, 999)"), 10)
```

### Exercise 4.2

Construct a query that retrieves all columns of rows from the `payment` table where the amount is greater than 500.

```{r}
head(dbGetQuery(con, 
           "SELECT *
           FROM payment
           WHERE amount > 500"), 10)
```

### Exercise 4.3

Construct a query that retrieves all columns of rows from the `payment` table where the amount is greater than 500 and less than 800.

```{r}
head(dbGetQuery(con, 
           "SELECT *
           FROM payment
           WHERE amount > 500 AND amount < 800"), 10)
```

## Cleanup

Run the following chunk to disconnect from the connection.

```{r}
# clean up
dbDisconnect(con)
```
