---
title: "BIOSTAT 620 HW 4"
author: "Kyoka Ono"
format: html
editor: visual
embed-resources: true
---

## HPC

For Questions 1-2, rewrite the provided R functions to make them faster. It is OK (and recommended) to take a look at StackOverflow and Google.

Using the synthetic dataset provided, show that your new function returns the same output as the original function. Then use `microbenchmark` to show that your version is faster and provide a visualization of the runtimes.

```{r}
library(microbenchmark)
library(parallel)
library(RSQLite)
library(DBI)
library(plotly)
library(ggplot2)
library(dplyr)
```

### Question 1 (6 points)

-   `fun2`: Calculate the running (cumulative) total along each row of a numeric matrix. Output should be a matrix with dimensions equal to the input matrix.

```{r}
# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j -  1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  t(apply(mat, 1, cumsum))
}
```

Generate and test using the following synthetic data (`dat`):

```{r}
# Use the functions with this data
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
```

```{r}
fun2(dat)[1:5,1:5]
```

```{r}
fun2alt(dat)[1:5,1:5]
```

```{r}
microbenchmark(fun2(dat), fun2alt(dat))
```

### Question 2 (8 points)

-   `fun3`: Calculate the running (cumulative) total of a vector, where the sum resets when a zero is encountered.

```{r}
fun3 <- function(x) {
  out <- numeric(length(x))
  cumsum_x <- 0
  for (i in seq_along(x)) {
    if (x[i] == 0) {
      cumsum_x <- 0
    } else {
      cumsum_x <- cumsum_x + x[i] 
    }
    out[i] <- cumsum_x
  }
  out
}
```

Generate your own synthetic datasets to test accuracy and speed.

```{r}
# testing accuracy 
test_data <- c(1,3,4,
                5,7,0,
                3,7,4,
                4,2,7,
                9,4,0)

fun3(test_data)
```

```{r}
# testing speed
system.time(fun3(test_data))
```

### Question 3 (12 points)

For this question, we will ask you to parallelize a process for computationally estimating the value of (pi).

The following function allows us to estimate $\pi$ through a simple simulation:

```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

In order to get more accurate estimates, we can run this function multiple times and take the mean:

```{r}
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

Rewrite the previous code using `parLapply()` (or your parallelization method of choice) to parallelize it. Run the code once, using `system.time()`, to show that your version is faster.

Visualize your results in two ways:

1.  Show that both versions produce similar ranges of estimates of $\pi$.
2.  Show that the parallel versions runs faster.

```{r}
# YOUR CODE HERE
system.time({
  cl <- makeCluster(7L)
  clusterExport(cl, "sim_pi")
  
  ans <- parLapply(cl = cl,
                   X = 1:4000,
                   fun = function(i){
                     sim_pi(1000)
                   })
 
  stopCluster(cl)
  
  ans <- do.call(rbind, ans)
  
  print(mean(ans))
})
```

## SQL

Dataset Description

The data we will be using is a time-series for a small mammal community in southern Arizona. This is part of a project studying the effects of rodents and ants on the plant community that has been running for almost 40 years. The rodents are sampled on a series of 24 plots, with different experimental manipulations controlling which rodents are allowed to access which plots.

Download data

You'll need the following three files from this repository:

-   `survey.csv`

-   `species.csv`

-   `plots.csv`

Setup a temporary database of ecology data by editing and running the following code chunk:

```{r}
# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Read in tables
survey <- read.csv("/Users/kyokaono/Desktop/BIOSTAT620W26/surveys.csv")
species <- read.csv("/Users/kyokaono/Desktop/BIOSTAT620W26/species.csv")
plots <- read.csv("/Users/kyokaono/Desktop/BIOSTAT620W26/plots.csv")

# Copy data.frames to database
dbWriteTable(con, "survey", survey)
dbWriteTable(con, "species", species)
dbWriteTable(con, "plots", plots)
```

When you write a new chunk, remember to replace the `r` with `sql`, `connection=con`. Some of these questions will require you to use an inner join. Read more about them here

<https://www.w3schools.com/sql/sql_join_inner.asp>

Answer each of the questions below with SQL queries. Make sure to show the results of your queries.

### Question 4 (6 points)

First, using the approach from Lab 9, create a query to see the structure of the surveys table. Then, using the surveys table, write a query to display the three date fields, species_id, and weight in kilograms (rounded to two decimal places), for individuals captured in 1999, ordered alphabetically by the species_id.

```{r}
dbGetQuery(con,
           "PRAGMA table_info(survey)")
```

```{r}
head(dbGetQuery(con,
           "SELECT year, month, day, species_id, ROUND(weight / 1000, 2) AS weight_kg
           FROM survey
           WHERE year = 1999
           ORDER BY species_id"))
```

### Question 5 (6 points)

Write a query that returns: the total weight, average weight, minimum and maximum weights for all animals caught over the duration of the survey. Can you modify it so that it outputs these values only for weights between 5 and 10 inclusive?

```{r}
dbGetQuery(con,
           "SELECT SUM(weight) AS total_weight, 
           AVG(weight) AS avg_weight, 
           MIN(weight) AS min_weight, 
           MAX(weight) AS max_weight
           FROM survey")
```

```{r}
dbGetQuery(con,
           "SELECT SUM(weight) AS total_weight, 
           AVG(weight) AS avg_weight, 
           MIN(weight) AS min_weight, 
           MAX(weight) AS max_weight
           FROM survey
           WHERE weight >= 5 AND weight <= 10")
```

### Question 6 (8 points)

Write a single query that returns:

1.  How many were counted each year, for each different species
2.  The average weights of each species in each year

Show the result and also store the resulting SQL data table as an R object. Hint, use the R function `DBI::dbGetQuery`. Note: to do so, you may need to, rather than running a SQL chunk, call SQL within an R chunk.

```{r}
# 1. How many were counted each year, for each different species
head(dbGetQuery(con,
           "SELECT species_id, year, COUNT(species_id) AS count, weight, AVG(weight) AS ave_weight
           FROM survey
           GROUP BY species_id, year"))
```

### Question 7 (8 points)

Write a query that returns, from the species table, the number of species in each taxa, only for the taxa with more than 10 species.

```{r}
dbGetQuery(con,
"SELECT taxa, COUNT(*) AS taxa_count
FROM species
GROUP BY taxa
HAVING taxa_count > 10")
```

### Question 8 (10 points)

Write a query that returns the number of animals caught of each genus in each plot. Order the results by plot number (ascending) and by descending number of individuals in each plot. Show the result and also store the resulting SQL data table as an R object.

```{r}
dim(survey)
dim(species)
dim(plots)
length(unique(survey$species_id))
length(unique(species$species_id))
```

```{r}
merged_data <- dbGetQuery(con,
"SELECT a.*, b.genus, b.species, b.taxa, c.plot_type
FROM survey AS a LEFT JOIN species AS b
ON a.species_id = b.species_id
LEFT JOIN plots as c
ON a.plot_id = c.plot_id")

dim(merged_data)

dbWriteTable(con, "merged_data", merged_data)
```

```{r}
head(dbGetQuery(con,
"SELECT plot_id, plot_type, genus, COUNT(*) AS count
FROM merged_data
GROUP BY genus, plot_id
ORDER BY plot_id ASC, count DESC"))
```

### Question 9 (8 points)

How many specimens of each species were captured in each type of plot, excluding specimens of unknown species?

```{r}
head(dbGetQuery(con,
"SELECT plot_type, species, COUNT(species) AS species_count
FROM merged_data
WHERE species IS NOT NULL
GROUP BY species, plot_type"))
```

### Question 10 (8 points)

Write a query that finds the average weight of each rodent species (i.e., only include species with Rodent in the taxa field).

```{r}
dbGetQuery(con,
"SELECT species, AVG(weight) AS weight_avg
FROM merged_data
WHERE taxa = 'Rodent'
GROUP BY species")
```

### Question 11 (10 points)

Using `plotly::ggplotly` and the data table resultant from question 8, make an interactive grouped bar chart of the number of genus by plot id, subsetting to plot id 1-5.

```{r}
data <- dbGetQuery(con,
"SELECT plot_id, plot_type, genus, COUNT(*) AS count
FROM merged_data
GROUP BY genus, plot_type
ORDER BY plot_id ASC, count DESC")

dbWriteTable(con, "data", data)
```

```{r}
data <- data %>% filter(plot_id %in% 1:5)
p <- ggplot(data = data, mapping = aes(x = factor(plot_id), y = count, fill = genus)) +
  geom_col(position = "dodge") + 
  labs(x = "Plot ID", y = "Count")  
  labs(title = "Number of Genus by Plot ID") + 
  theme(plot.title = element_text(hjust = 0.5))
ggplotly(p)
```

### Question 12 (10 points)

Using native `plotly` functions (no `ggplot2`) and the data table resultant from question 6, make an interactive line chart of average species weight over time for each species. Make sure to properly handle missing data. Can you do anything to reduce the noise in your plot?

```{r}
data_2 <- dbGetQuery(con,
           "SELECT species_id, year, COUNT(species_id) AS count, weight, AVG(weight) AS ave_weight
           FROM survey
           GROUP BY species_id, year")

dbWriteTable(con, "data_2", data_2)
```

```{r}
plot_ly(data_2, 
       x = ~year,
       y = ~ave_weight, 
       color = ~species_id,
       type = "scatter",
       mode = "lines", 
       hoverinfo = "text",
       text = ~paste(paste(species_id, ":", sep = ""),
                     paste("Year: ", year, sep = ""),
                     paste(" Average Weight: ", ave_weight, sep = ""),
                           sep = "<br>"))
```
