---
title: "BIOSTAT 620 Lab 8"
author: "Kyoka Ono"
format: html
editor: visual
embed-resources: true
---

### Problem 1: Vectorization

```{r, include = FALSE}
library(microbenchmark)
library(Rfast)
library(parallel)
```

1.  This function generates an `n x k` dataset with all its entries drawn from a Poisson distribution with mean `lambda`.

rpois(4,4) generates random variables

```{r}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n){
    x <- rbind(x, rpois(k, lambda))    
  }
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  x <- matrix(data = rpois(n,4), nrow = 100, ncol = 4, byrow = TRUE)
  return(x)
}
```

Show that `fun1alt` generates a matrix with the same dimensions as `fun1` and that the values inside the two matrices follow similar distributions. Then check, using `microbenchmark::microbenchmark`, the speed of the two functions:

```{r}
# Benchmarking
dim(fun1())
dim(fun1alt())
```

```{r}
head(fun1())
```

```{r}
head(fun1alt())
```

```{r}
microbenchmark(fun1(), fun1alt())
```

2.  This function finds the maximum value of each column of a matrix

```{r}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  colMaxs(x, value = TRUE)
}
```

Show that both functions return the same output for a given input matrix, x. Then check the speed of the two functions, as you did above.

```{r}
head(fun2(x))
```

```{r}
head(fun2alt(x))
```

```{r}
microbenchmark(fun2(x), fun2alt(x))
```

### Problem 2: Parallelization

The main assumption is that we can approximate the results of many repeated experiments by resampling observations from our original dataset, which reflects the population.

1.  This function implements a serial version of the bootstrap. Edit this function to parallelize the `lapply` loop, using whichever method you prefer. Rather than specifying the number of cores to use, use the number given by the `ncpus` argument, so that we can test it with different numbers of cores later.

```{r}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
  
  # THIS FUNCTION NEEDS TO BE PARALELLIZED
  # EDIT THIS CODE:
  cl <- makeCluster(ncpus)
  clusterExport(cl, c("dat","stat"), envir = environment())
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  stopCluster(cl)
  
  # Converting the list into a matrix
  ans <- do.call(rbind, ans)

  return(ans)
}
```

2.  Once you have a version of the `my_boot()` function that runs on multiple cores, check that it provides accurate results by comparing it to a parametric model:

```{r}
# Bootstrap of an OLS
my_stat <- function(d) coef(lm(y ~ x, data=d))

# DATA SIM
set.seed(1)
n <- 500; R <- 1e4

x <- cbind(rnorm(n)); y <- x*5 + rnorm(n)

# Checking if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)

# You should get something like this
t(apply(ans1, 2, quantile, c(.025,.975)))
##                   2.5%      97.5%
## (Intercept) -0.1372435 0.05074397
## x            4.8680977 5.04539763
ans0
##                  2.5 %     97.5 %
## (Intercept) -0.1379033 0.04797344
## x            4.8650100 5.04883353
```

3.  Check whether your version actually goes faster when it's run on multiple cores (since this might take a little while to run, we'll use `system.time` and just run each version once, rather than `microbenchmark`, which would run each version 100 times, by default).

```{r}
system.time(my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 1L) )
```

```{r}
system.time(my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L) )
```

```{r}
system.time(my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 3L) )
```

Yes, it goes faster when it is run on multiple cores.

### Problem 3: Short Answer

1.  Explain in your own words how you can decide if a problem is fit for parallelization?

If the code is already vectorized and if the computation can be done simultaneously to different inputs, especially when there is a large amount of data that needs to go through the same computation.

2.  We have covered vectorization and parallelization in this lab. Explain in your own words how you can think about the relationship between vectorization and parallelization? How do we know whether we should use one or the other or both?

A code should be vectorized first. If the data is large, it should be parallelized so that the task can be split across multiple cores.
